---
title: "Augmenting small training sets for machine learning"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Augmenting small training sets for machine learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
ok <- requireNamespace("rpart", quietly = TRUE) && requireNamespace("MASS", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 72, out.width = "100%", eval = ok)
```

Machine-learning models often need more data than we have. A synthesizer
fitted to a small sample can generate as many additional training records
as needed. This vignette shows how to do that with `augment_data()` and,
just as importantly, how to check whether it actually helps.

```{r}
library(SynthesizerPlus)
library(rpart)
```

## What augmentation can and cannot do

A synthesizer only knows what is in its training sample, so synthetic data
**add no new information**. What they provide is a *smoothed* version of the
training distribution: new combinations of values that are plausible under
the fitted marginals and copula. For flexible, high-variance learners
(regression trees, nearest neighbours, boosting, neural networks) trained on
a small sample this acts as a regulariser and can reduce the prediction
error. For a model that is already smooth, such as linear regression, there
is little to gain.

Whether it helps must therefore be **measured**, following four rules:

1. split the real data into a training and a test set first;
2. fit the synthesizer on the training set only;
3. evaluate on the held-out *real* test set (never on synthetic data);
4. repeat over several random splits.

## Tabular data: 60 observations

We pretend that only 60 records of the Boston housing data are available
and predict the median house value `medv`.

```{r}
data(Boston, package = "MASS")
rmse <- function(pred, obs) sqrt(mean((pred - obs)^2))
tree <- function(f, d) rpart(f, d, control = rpart.control(cp = 0.005, minsplit = 5))

one_split <- function(seed) {
  set.seed(seed)
  idx <- sample(nrow(Boston), 60)
  train <- Boston[idx, ]
  test <- Boston[-idx, ]
  big <- augment_data(train, n = 2000, seed = seed)   # 60 real + 2000 synthetic
  c(tree_real = rmse(predict(tree(medv ~ ., train), test), test$medv),
    tree_augmented = rmse(predict(tree(medv ~ ., big), test), test$medv),
    lm_real = rmse(predict(lm(medv ~ ., train), test), test$medv),
    lm_augmented = rmse(predict(lm(medv ~ ., big), test), test$medv))
}
res <- t(sapply(1:6, one_split))
round(colMeans(res), 2)
mean(res[, "tree_augmented"] < res[, "tree_real"])
```

The regression tree improves clearly and consistently; the linear model does
not. Before training, it is worth checking that the synthetic training data
resemble the real training data:

```{r, fig.height = 4.5}
set.seed(1)
train <- Boston[sample(nrow(Boston), 60), ]
syn <- generate(fit_synthesizer(train), n = 2000, seed = 1)
plot_marginals(train, syn, vars = c("medv", "lstat", "rm", "crim"))
```

`augment_data()` accepts all arguments of `fit_synthesizer()`, e.g. `by =`
to stratify on a categorical variable, and `label =` to add a column marking
the synthetic rows (useful for down-weighting them in training).

## Time series: forecasting from a short history

For forecasting, an ML model is trained on windows of lagged values. With one
short series there are few windows; `augment_data()` on a `ts` object returns
the real series followed by `nsim` synthetic series, and `ts_windows()`
stacks the windows of all of them.

We use the first 48 months of `ldeaths` for training and forecast the last
24 months one step ahead.

```{r}
lags <- 12
n_train <- 48
train_ts <- window(ldeaths, end = time(ldeaths)[n_train])
all_w <- ts_windows(ldeaths, lags = lags)
test_w <- all_w[(n_train - lags + 1):nrow(all_w), ]   # targets in the test period

forecast_rmse <- function(seed) {
  series <- augment_data(train_ts, nsim = 50, order = 12, seed = seed)
  real_w <- ts_windows(series[1], lags = lags)
  aug_w <- ts_windows(series, lags = lags)
  real_w$series <- aug_w$series <- NULL
  c(tree_real = rmse(predict(tree(y ~ ., real_w), test_w), test_w$y),
    tree_augmented = rmse(predict(tree(y ~ ., aug_w), test_w), test_w$y),
    lm_real = rmse(predict(lm(y ~ ., real_w), test_w), test_w$y))
}
round(colMeans(t(sapply(1:6, forecast_rmse))), 1)
```

Two lessons:

* Augmentation again helps the tree, but the linear autoregression on the
  real data is still the best forecaster here. Synthetic data make a
  flexible model better, not necessarily better than a well-specified simple
  one.
* The synthesizer must reproduce the structure the forecaster relies on.
  For monthly data with a short history, AIC selects a VAR order below the
  seasonal period, and the synthetic series then lack the yearly cycle;
  fixing `order = 12` solves this. Always compare the autocorrelation of
  real and synthetic series:

```{r}
fit_ts <- fit_synthesizer(train_ts, order = 12)
plot_acf(train_ts, generate(fit_ts, n = 2000, seed = 3), lag_max = 24)
```

## Checklist

* Fit the synthesizer on training data only; evaluate on real test data.
* Compare at least: real only, real + synthetic, and a simple baseline.
* Inspect the synthetic training data (`compare_synthetic()`,
  `plot_marginals()`, `plot_acf()`) before trusting the results.
* Tune the amount of synthetic data and the synthesizer settings
  (`by`, `copula`, `order`) with cross-validation on the training data,
  not on the test set.
